---
title: "Introduction to Unix Command-line - Part 2"
author: "Natalie Elphick"
date: "March 12th 2024"
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../docs"
    )
  })
output:
  revealjs::revealjs_presentation:
    css: style.css
---
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = './unix_workshop_2024')
```

## 

<center>*Press the ? key for tips on navigating these slides*</center>


## Introductions

**Natalie Elphick**   
Bioinformatician I    

<br>


**Yihang Xin (TA)**    
Software Engineer II  

<br>

# Setup

Run the following commands if you did not attend part 1:   


```{r, engine='bash', eval=FALSE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
mkdir unix_workshop
```


```{r, engine='bash', eval=FALSE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
cd unix_workshop
```


```{r, engine='bash', eval=FALSE, results="hide", highlight=FALSE, comment=NA, echo = TRUE}
curl -L -o unix_workshop_2024.tar.gz 'https://www.dropbox.com/s/smb12au2y82jmvq/unix_workshop_2024.tar.gz?dl=0'
```


```{r, engine='bash', eval=FALSE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
tar -xzf unix_workshop_2024.tar.gz
```


```{r, engine='bash', eval=FALSE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
cd unix_workshop_2024
```

```{r, engine='bash', eval=TRUE, results="hide", highlight=FALSE, comment=NA, echo = TRUE}
curl -o part_2/homo_sapiens.refseq.tsv.gz https://ftp.ensembl.org/pub/current_tsv/homo_sapiens/Homo_sapiens.GRCh38.111.refseq.tsv.gz
```

# File Compression

## Command-line tools for compression

- Compression reduces the size of a file
- `gzip` : compresses a file and replaces it with a compressed version (.gz)
- `tar` : create and manipulate archive files

<br>

**Archive**: a single file that contains one or more files and/or folders that have been compressed



## gzip/gunzip: compress/uncompress a file


```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
gunzip part_2/homo_sapiens.refseq.tsv.gz
du -h part_2/homo_sapiens.refseq.tsv
```

- The uncompressed file is 27 megabytes

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
gzip part_2/homo_sapiens.refseq.tsv
du -h part_2/homo_sapiens.refseq.tsv.gz
```

- Compressing it makes it a 10th of the size

## Note

- The magnitude of the compression depends on type of data
- The units for file sizes are not the same across all systems
  - Some systems define a kilobyte as 1000 bytes, while others define it as 1024 bytes



## tar: compressing folders into archives

- Does not provide compression on its own, it uses gzip to create compressed archive files

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
tar -czf part_1.tar.gz part_1
ls -l
```

- -c: create a new archive
- -f: specify the name of the archive file
- -z: compress the archive with gzip

## Unarchiving

- We did this in part 1 to unarchive the workshop folders


```{r, engine='bash', eval=FALSE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
tar -xzf part_1.tar.gz
```
- -x: extract an archive
- -z: uncompress the archive with gzip
- -f: specify the name of the archive file


## gunzip -c: cat compressed files

- To avoid uncompressing a large file just to read its contents, we can use `gunzip -c`
- This will output the the file to the terminal


```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
gunzip -c part_2/homo_sapiens.refseq.tsv.gz | head
```



# System Variables

## What are system variables?

- Special variables that contain information about the system's configuration and state
- Used by the OS and programs to change their behavior based on the system's state

Example:
```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
echo $HOME
```


## Common System Variables

- **$PWD** : The working directory
- **$HOME** : The current user's home directory
- **$PS1** : the shell prompt string
- **$TEMP** : location of temporary files



## PATH:  locations of executable files

- When you enter a command, the OS searches the directories in the `$PATH` to find its associated executable file

```{r, engine='bash', eval=FALSE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
echo $PATH
```


```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = FALSE}
echo "/usr/local/bin:/usr/bin:/bin:/usr/sbin:/sbin:/usr/local/go/bin:/usr/local/mysql/bin"
```



- The OS will check these directories in the order they appear and use the first executable it finds

## export: set system variables

- Useful for setting variables you want to be used across programs
- You can add new software to your `$PATH` like this:

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = FALSE}
echo 'export PATH="/path/to/new/software:$PATH"'
```

- This will modify the `$PATH` for the current terminal session

## Modifying the PATH for all future terminal sessions

- Add the export line to your `~/.bashrc` or `~/.zshrc`
- **Proceed with caution** 
  - Make backups of these and read this [guide](https://tldp.org/LDP/Bash-Beginners-Guide/html/sect_03_01.html)
  - Changing the `$PATH` incorrectly can <span style="color:#b01212;font-weight:bold">break system functionality</span>


## which: locate the executable associated with a command

- This command shows the location of the executable that the OS finds


```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
which ls
```

- Useful to check if there are multiple versions of a software installed

# Shell Scripting


## What is a script?

- Scripts are executable files for reusing code
- By convention scripts end in `.sh`
- This first line of the script is called the shebang

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = FALSE}
echo "nano part_2/example_script.sh"
cat ../materials/example_script.sh > part_2/example_script.sh
```

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = FALSE}
cat ../materials/example_script.sh | head -n 1
```

- The text that follows `#!` tells the OS where the interpreter is 

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
which bash
```

## chmod: making a script executable

- By default, files are not executable

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
ls -l part_2/example_script.sh
```

- We can set the execute bit like this

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
chmod u+x part_2/example_script.sh
ls -l part_2/example_script.sh
```

## Example

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = FALSE}
cat ../materials/example_script.sh
```

## Let's run it

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
./part_2/example_script.sh part_2/homo_sapiens.refseq.tsv.gz
```

## Loops

- Useful for iterating over lines of a file or lists

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
for i in {1..3}
do

echo $i

done
```

## While loops

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
count=0

while [ $count -lt 5 ]        # loop while count is less than 5
do
    echo $count
    count=$((count+1))
done
```


## If statements

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
x=5

if [ $x -gt 10 ]                      # check if x is greater than 10
then
    echo "x is greater than 10"
else
    echo "x is not greater than 10"
fi                                    # end if statement
```

# Other Useful Commands

## sed : stream editor

- Parses and transforms text, using a compact programming language
- It reads and modifies text line by line from a file or input stream
- Supports [regular expressions](https://v4.software-carpentry.org/regexp/index.html)
- Useful for replacing text

Example:
```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = FALSE}
echo "sed 's/search_string/replace_string/g' input.txt > output.txt"
```

## ssh : secure shell - conect to remote server

- Logging in to a remote server
- Remote desktop for the terminal
```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = FALSE}
echo "ssh username@remote"
```
- The `username` would be your user on the remote server and `remote` is the hostname or IP address of the remote server or computer

## scp :  secure copy

- Copy files from a remote server or computer

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = FALSE}
echo "scp [options] [source] [destination]"
```
- Copy from local to remote

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = FALSE}
echo "scp /path/to/local/file.txt username@remote:/path/to/remote/directory/"
```
- Copy from remote to local

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = FALSE}
echo "scp username@remote:/path/to/file.txt /path/to/local/directory/"
```

- -r : copy a whole folder

# AWK

## awk : processing structured data

- A small programming language that is designed to work with structured data
- Has more complicated syntax but is faster at processing large files
- Designed to read a file or input stream line by line
- Operates on **records** (lines) and **fields** (columns)

Basic command:

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = FALSE}
echo "awk options 'pattern {action}' input_file"
```

## Example : Sum the first 2 columns of a file

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
awk -F '\t' '{print $1+$2}' part_1/list_numbers.tsv
```

- -F : provides the field separator
- `$1,$2` : the first and second fields

## Example : Find the average of a column

- For this example we only want the average if the 5th column equals "RefSeq_mRNA" 

```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo = TRUE}
gunzip -c part_2/homo_sapiens.refseq.tsv.gz | \
awk -F '\t' '$5 == "RefSeq_mRNA" {sum += $7; count++} \
END {print sum / count}'

```

## Resources for learning AWK and sed

- [The GNU AWK manual](https://www.gnu.org/software/gawk/manual/gawk.html)
- [AWK Tutorial by Bruce Barnett](https://www.grymoire.com/Unix/Awk.html)
- [Sed Tutorial by Bruce Barnett](https://www.grymoire.com/Unix/Sed.html)



# End of Part 2




## Survey

- Please take some time to fill out the workshop survey:
https://www.surveymonkey.com/r/DY7K5ZY


## Additional learning materials

- Software carpentry provides a self paced course: 
  - [The Unix Shell](https://swcarpentry.github.io/shell-novice/)

- Free online books:
  - [The Unix Workbench](https://seankross.com/the-unix-workbench/index.html)
  - [The Linux Command Line](http://linuxcommand.org/tlcl.php)




## Upcoming Data Science Training Program Workshops

[Introduction to Pathway Analysis](https://gladstone.org/events/introduction-pathway-analysis-1)   
April 2, 2024 1:00-4:00pm PDT

[Statistics of Enrichment Analysis Methods](https://gladstone.org/events/statistics-enrichment-analysis-methods-0)  
April 11-April 12, 2024 1:00-3:00pm PDT

[Working on Wynton](https://gladstone.org/events/working-wynton)   
April 15, 2024 1:00-4:00pm PDT

[Introduction to Linear Mixed Effects Models](https://gladstone.org/events/introduction-linear-mixed-effects-models)  
April 25-April 26, 2024 1:00-3:00pm PDT



[Complete Schedule](https://gladstone.org/events?series=data-science-training-program)




```{r, engine='bash', eval=TRUE, results='markup', highlight=FALSE, comment=NA, echo=FALSE}
rm part_2/example_script.sh
rm part_2/homo_sapiens.refseq.tsv*
rm part_1.tar.gz
```


