---
title: "Introduction to Unix Command-line"
subtitle: "Part 2"
author: "Natalie Gill"
institute: "Gladstone Bioinformatics Core"
date: "March 10, 2025"
date-format: MMMM DD, YYYY
format:
  revealjs:
    theme: unix-theme.scss
    highlight-style: arrow
    code-line-numbers: false
    template-partials:
      - title-slide.html
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(comment = "")
knitr::opts_knit$set(root.dir = './unix_workshop')
```

##

<center>*Press the ? key for tips on navigating these slides*</center>


## Introductions

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;**Natalie Gill**   
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;*Bioinformatician II*

<br>


# Setup

- If you did not attend part 1
- Copy and paste each of these into your terminal, press enter after each one:

```{bash}
#| eval: false
#| echo: true

mkdir unix_workshop
```


```{bash}
#| eval: false
#| echo: true

cd unix_workshop
```

```{bash}
#| eval: false
#| echo: true

curl -L -o unix_workshop.tar.gz 'https://www.dropbox.com/scl/fi/tdzpoivf7mienlenunqhf/unix_workshop.tar.gz?rlkey=6bfxnqgc5n4lgc9mc80ld75z4&dl=0'
```


```{bash}
#| eval: false
#| echo: true

tar -xzf unix_workshop.tar.gz
```


```{bash}
#| eval: false
#| echo: true

cd unix_workshop
```

```{bash}
#| eval: true
#| echo: false
#| output: false
curl -o part_2/homo_sapiens.refseq.tsv.gz https://ftp.ensembl.org/pub/current_tsv/homo_sapiens/Homo_sapiens.GRCh38.115.refseq.tsv.gz
```


# File Compression

## Command-line tools for compression

- Compression reduces the size of a file
- `gzip` : compresses a file and replaces it with a compressed version (.gz)
- `tar` : create and manipulate archive files

<br>

**Archive**: a single file that contains one or more files and/or folders that have been compressed



## gzip/gunzip: compress/uncompress a file


```{bash}
#| echo: true
gunzip part_2/homo_sapiens.refseq.tsv.gz
du -h part_2/homo_sapiens.refseq.tsv
```

- The uncompressed file is 33 megabytes

```{bash}
#| echo: true
gzip part_2/homo_sapiens.refseq.tsv
du -h part_2/homo_sapiens.refseq.tsv.gz
```

- Compressing it makes it a 10th of the size

## Note

- The magnitude of the compression depends on type of data
- The units for file sizes are not the same across all systems
  - Some systems define a kilobyte as 1000 bytes, while others define it as 1024 bytes



## tar: compressing folders into archives

- Does not provide compression on its own, it uses gzip to create compressed archive files

```{bash}
#| echo: true
tar -czf part_1.tar.gz part_1
ls -l
```

- -c: create a new archive
- -f: specify the name of the archive file
- -z: compress the archive with gzip

## Unarchiving

- We did this in part 1 to unarchive the workshop folders


```{bash}
#| eval: false
#| echo: true
tar -xzf part_1.tar.gz
```
- -x: extract an archive
- -z: uncompress the archive with gzip
- -f: specify the name of the archive file


## gunzip -c: cat compressed files

- To avoid uncompressing a large file just to read its contents, we can use `gunzip -c`
- This will output the the file to the terminal


```{bash}
#| echo: true
gunzip -c part_2/homo_sapiens.refseq.tsv.gz | head
```



# System Variables

## What are system variables?

- Special variables that contain information about the system's configuration and state
- Used by the OS and programs to change their behavior based on the system's state

Example:
```{bash}
#| echo: true
echo $HOME
```


## Common System Variables

- **$PWD** : The working directory
- **$HOME** : The current user's home directory
- **$PS1** : the shell prompt string
- **$TMPDIR** : location of temporary files

These can change depending on the specific OS or program, **TMPDIR** can also be **TEMP**, **TEMPDIR** and **TMP**.



## PATH:  locations of executable files

- When you enter a command, the OS searches the directories in the `$PATH` to find its associated executable file

```{bash}
#| eval: false
#| echo: true
echo $PATH
```

```
/usr/local/bin:/usr/bin:/bin:/usr/sbin:/sbin:/usr/local/go/bin:/usr/local/mysql/bin
```



- The OS will check these directories in the order they appear and use the first executable it finds

## export: set system variables

- Useful for setting variables you want to be used across programs
- You can add new software to your `$PATH` like this:

```{bash}
#| eval: false
#| echo: true
export PATH="/path/to/new/software:$PATH"
```

- This will modify the `$PATH` for the current terminal session

## Modifying the PATH for all future terminal sessions

- Add the export line to your `~/.bashrc` or `~/.zshrc`
- **Proceed with caution**
  - Make backups of these and read this [guide](https://tldp.org/LDP/Bash-Beginners-Guide/html/sect_03_01.html)
  - Changing the `$PATH` incorrectly can <span style="color:#b01212;font-weight:bold">break system functionality</span>


## which: locate the executable associated with a command

- This command shows the location of the executable that the OS finds


```{bash}
#| echo: true
which ls
```

- Useful to check if there are multiple versions of a software installed

# Shell Scripting


## What is a script?

- Scripts are executable files for reusing code
- By convention scripts end in `.sh`
- This first line of the script is called the shebang

```{bash}
#| eval: false
#| echo: true
nano part_2/example_script.sh
```

```{bash}
#| echo: false
cat ../materials/example_script.sh > part_2/example_script.sh
```

```{bash}
#| echo: false
cat ../materials/example_script.sh | head -n 1
```

- The text that follows `#!` tells the OS where the interpreter is

```{bash}
#| echo: true
which bash
```

## chmod: making a script executable

- By default, files are not executable

```{bash}
#| echo: true
ls -l part_2/example_script.sh
```

- We can set the execute bit like this

```{bash}
#| echo: true
chmod u+x part_2/example_script.sh
ls -l part_2/example_script.sh
```

## Example

```{r}
#| echo: false
#| output: asis
cat("```bash\n")
cat(paste(readLines("../materials/example_script.sh"), collapse = "\n"))
cat("\n```\n")
```

## Let's run it

```{bash}
#| echo: true
./part_2/example_script.sh part_2/homo_sapiens.refseq.tsv.gz
```

## Loops

- Useful for iterating over lines of a file or lists

```{bash}
#| echo: true
for i in {1..3}
do

echo $i

done
```

## While loops

```{bash}
#| echo: true
count=0

while [ $count -lt 5 ]        # loop while count is less than 5
do
    echo $count
    count=$((count+1))
done
```


## If statements

```{bash}
#| echo: true
x=5

if [ $x -gt 10 ]                      # check if x is greater than 10
then
    echo "x is greater than 10"
else
    echo "x is not greater than 10"
fi                                    # end if statement
```

# Other Useful Commands

## sed : stream editor

- Parses and transforms text, using a compact programming language
- It reads and modifies text line by line from a file or input stream
- Supports [regular expressions](https://regexone.com/lesson/introduction_abcs)
- Useful for replacing text

Example:
```{bash}
#| echo: true
#| eval: false
sed 's/search_string/replace_string/g' input.txt > output.txt
```

## ssh : secure shell - conect to remote server

- Logging in to a remote server
- Remote desktop for the terminal
```{bash}
#| echo: true
#| eval: false
ssh username@remote
```
- The `username` would be your user on the remote server and `remote` is the hostname or IP address of the remote server or computer

## scp :  secure copy

- Copy files from a remote server or computer

```{bash}
#| echo: true
#| eval: false
scp [options] [source] [destination]
```
- Copy from local to remote

```{bash}
#| echo: true
#| eval: false
scp /path/to/local/file.txt username@remote:/path/to/remote/directory/
```
- Copy from remote to local

```{bash}
#| echo: true
#| eval: false
scp username@remote:/path/to/file.txt /path/to/local/directory/
```

- -r : copy a whole folder

# AWK

## awk : processing structured data

- A small programming language that is designed to work with structured data
- Has more complicated syntax but is faster at processing large files
- Designed to read a file or input stream line by line
- Operates on **records** (lines) and **fields** (columns)

Basic command:

```{bash}
#| echo: true
#| eval: false
awk options 'pattern {action}' input_file
```

## Example : Sum the first 2 columns of a file

```{bash}
#| echo: true
 awk -F '\t' '{print $1+$2}' part_1/list_numbers.tsv
```

- -F : provides the field separator
- `$1,$2` : the first and second fields

## Example : Find the average of a column

- For this example we only want the average if the 5th column equals "RefSeq_mRNA"

```{bash}
#| echo: true
gunzip -c part_2/homo_sapiens.refseq.tsv.gz | \
awk -F '\t' '$5 == "RefSeq_mRNA" {sum += $7; count++} \
END {print sum / count}'

```

## Resources for learning AWK and sed

- [The GNU AWK manual](https://www.gnu.org/software/gawk/manual/gawk.html)
- [AWK Tutorial by Bruce Barnett](https://www.grymoire.com/Unix/Awk.html)
- [Sed Tutorial by Bruce Barnett](https://www.grymoire.com/Unix/Sed.html)



# End of Part 2



## Notes on Using AI for UNIX CLI {.smaller}

- AI tools (ChatGPT, Claude, Gemini) can help generate, explain, and debug Unix commands
- Always verify AI-generated commands before running them, especially destructive ones (rm, mv)
- AI can hallucinate flags or options — confirm with `man` or `--help`
- Provide context for better results: your shell, OS, and goal
- If you can't explain what a command does, don't run it
- See the Gladstone Bioinformatics Core position paper on responsible use of AI in bioinformatics: [link](https://www.dropbox.com/scl/fi/hyk0n0egz3rlxonuai6t7/Bioinformatics-Core-AI-Position-Paper-v1.0.pdf?rlkey=1gq3icconjym6jejm6aayrvup&dl=0)




## Additional learning materials


- Learn Regular Expressions:
  - [RegexOne](https://regexone.com/lesson/introduction_abcs)

- Software carpentry provides a self paced course:
  - [The Unix Shell](https://swcarpentry.github.io/shell-novice/)

- Free online books:
  - [The Unix Workbench](https://seankross.com/the-unix-workbench/index.html)
  - [The Linux Command Line](http://linuxcommand.org/tlcl.php)


## Workshop survey
- Please fill out our [workshop survey](https://www.surveymonkey.com/r/bioinfo-training26) so we can continue to improve these workshops


## Upcoming Data Science Training Program Workshops {.smaller}


**Working on Wynton**    
March 16-March 17, 2026 9:00am-12:00pm PDT    

**Introduction to Pathway Analysis**    
March 19, 2026 1:00-4:00pm PDT    

**Statistics of Enrichment Analysis Methods**    
March 23-March 24, 2026 1:00-3:00pm PDT    

**Introduction to R Data Analysis**    
April 6-April 7, 2026 1:00-4:00pm PDT    

**Single Cell RNA-Seq Analysis Using R**    
April 13-April 14, 2026 9:30am-12:00pm PDT    

**AI Fluency for Biologists**    
April 20-April 21, 2026 1:00-4:00pm PDT    





```{bash}
#| echo: false
rm part_2/example_script.sh
rm part_2/homo_sapiens.refseq.tsv*
rm part_1.tar.gz
```

