---
title: "Machine Learning Using R - Part 1"
subtitle: "Data Exploration & Preprocessing"
author: "Gladstone Bioinformatics Core"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

Welcome to the Machine Learning Using R workshop! In this first part, we'll explore our dataset and prepare it for machine learning algorithms.

## Why Data Preprocessing Matters

Before applying any machine learning algorithm, it's crucial to understand and prepare your data. Different ML algorithms have different requirements:

- **Missing data**: Most algorithms cannot handle missing values
- **Categorical variables**: May need to be encoded as numbers
- **Feature scaling**: Some algorithms are sensitive to the scale of features

**Key takeaway:** Always clean and preprocess your data before applying ML algorithms!

---

# Load Required Libraries

```{r libraries}
# Data manipulation & visualization (includes ggplot2)
library(tidyverse)
```

---

# Load and Explore the Data

We'll use a COVID-19 symptoms dataset containing patient symptoms and test results.  
- **Target variable**: `covid19_test_results` (what we want to predict)  
- **Features**: All other columns (symptoms and risk factors)

```{r load-explore-data}
# Load the data
covid_data <- read.csv("data/covid-symptoms.csv")

# Check target distribution
print(table(covid_data$covid19_test_results))
```  

---

# Handling Missing Data

Most ML algorithms cannot handle missing values. Two strategies:  
- **Remove rows**: Simple, but may lose data  
- **Impute values**: Replace with column mean, median, or mode  

## Why Is the Data Missing?

Which strategy is appropriate depends on **why** the values are missing. There are three types:

| Type | Meaning | Example |
|------|---------|---------|
| **MCAR** - Missing Completely At Random | Missingness has nothing to do with any variable, observed or not | A sample tube is dropped, or a sensor briefly glitches |
| **MAR** - Missing At Random | Missingness depends on *other observed* variables, not on the missing value itself | Older patients are more likely to have a lab test ordered, and age is recorded |
| **MNAR** - Missing Not At Random | Missingness depends on the *missing value itself* | Gene expression below a detection limit is reported as blank |

**Rule of thumb:**  
- **MCAR**: Removal or imputation are both reasonable (removal only costs sample size)  
- **MAR**: Imputation is usually better, since removing rows can bias your results  
- **MNAR**: Neither is safe on its own - you need to understand the *mechanism* (ask the people who collected the data!)  

## How Much Is Missing?

Before choosing a strategy, look at which columns have missing values and how many:

```{r missing-counts}
# Count missing (NA) or blank values in each column of the raw data
missing_counts <- sapply(covid_data, function(x) sum(is.na(x) | x == ""))

# Show only columns with missing values, largest first
missing_counts <- sort(missing_counts[missing_counts > 0], decreasing = TRUE)
data.frame(
  column = names(missing_counts),
  n_missing = missing_counts,
  pct_missing = round(100 * missing_counts / nrow(covid_data), 1),
  row.names = NULL
)
```

Thinking about *why* these might be missing (we can't know for sure from the data alone - which is exactly the challenge):  
- **`temperature`**: could be MAR - perhaps not measured for certain visit types  
- **`rhonchi`, `wheezes`** (lung exam findings): might be missing when the exam wasn't performed, for example on patients who looked well. That would be MNAR, because the missingness is related to the finding itself  
- **`cough_severity`**: blank for most patients - we'll treat blank as "No cough" below, which is an *assumption* about why it's missing  

**Going further:** The `naniar` package (`vis_miss()`, `mcar_test()`) can visualize missingness patterns and formally test for MCAR.

## Imputation

We'll use **median imputation** for numeric columns to keep this demo simple:

```{r imputation}
# Create a copy for imputation
covid_imputed <- covid_data

# Identify numeric columns (excluding target)
numeric_cols <- names(covid_imputed)[sapply(covid_imputed, is.numeric)]

# Impute numeric columns with median
for (col in numeric_cols) {
  if (sum(is.na(covid_imputed[[col]])) > 0) {
    median_val <- median(covid_imputed[[col]], na.rm = TRUE)
    covid_imputed[[col]][is.na(covid_imputed[[col]])] <- median_val
    cat("Imputed", col, "with median:", median_val, "\n")
  }
}

# Check if any missing values remain
cat("Remaining missing values:", sum(is.na(covid_imputed)))
```

---

# Handling Categorical Variables

Different algorithms have different requirements for categorical variables:  
- **K-means, SVM, XGBoost**: Require numeric input → need encoded version

We'll prepare both versions (original and encoded) so you can see the encoding step, even though both algorithms used in this workshop end up needing the encoded one.

### Handle Missing Values

First, let's handle missing values in the original categorical variable:

```{r handle-categorical-missing}
# Replace missing/empty values with "No cough"
covid_imputed$cough_severity <- ifelse(
  is.na(covid_imputed$cough_severity) | covid_imputed$cough_severity == "",
  "No cough",
  covid_imputed$cough_severity
)

# Convert to factor to ensure proper handling
covid_imputed$cough_severity <- as.factor(covid_imputed$cough_severity)

# Check the distribution
table(covid_imputed$cough_severity)
```

### Create Encoded Version

```{r label-encoding}
# Encode ordinal cough_severity: No cough=0 < Mild=1 < Moderate=2 < Severe=3
# This version will be used for K-means (Part 2)
covid_imputed$cough_severity_encoded <- case_when(
  covid_imputed$cough_severity == "No cough" ~ 0,
  covid_imputed$cough_severity == "Mild" ~ 1,
  covid_imputed$cough_severity == "Moderate" ~ 2,
  covid_imputed$cough_severity == "Severe" ~ 3,
  TRUE ~ 0
)
table(covid_imputed$cough_severity_encoded)
```

**Note:** We keep both `cough_severity` (original categorical) and `cough_severity_encoded` (numeric) in our dataset:  
- **Part 2 (K-means)**: Will use `cough_severity_encoded` (requires numeric input)  
- **Part 3 (XGBoost)**: Will also use `cough_severity_encoded` (XGBoost requires numeric input too)  

---

# Feature Scaling

Some algorithms (K-means, SVM, KNN) use distance calculations and are sensitive to feature scales.

```{r scale-demo}
# Compare scales - temperature vs binary symptoms
cat("Temperature range:", range(covid_imputed$temperature, na.rm = TRUE), "\n")
```

If temperature ranges from 36-40 and symptoms are 0-1, temperature will dominate distance calculations!

**Standardization (Z-score):** transforms features to mean=0, sd=1

```{r standardization}
# Example: Standardize temperature: (x - mean) / sd
temperature_scaled <- scale(covid_imputed$temperature)

cat("Original - Mean:", round(mean(covid_imputed$temperature), 2), 
    "SD:", round(sd(covid_imputed$temperature), 2), "\n",
    "Scaled - Mean:", round(mean(temperature_scaled), 4), 
    "SD:", round(sd(temperature_scaled), 4), "\n")
```

**Note:** We demonstrate scaling here, but we'll scale all features together in Part 2 when needed for K-means. XGBoost (Part 3) doesn't require scaling - like other tree-based methods, it only cares about the order of values, not their scale.

**Looking ahead:** Before scaling in Part 2, we'll also check for redundant (highly correlated) features, since they can skew K-means clustering.

---

# Prepare Final Dataset for ML

```{r final-dataset}
# Create the final clean dataset for machine learning
# Note: We include both cough_severity (categorical) and cough_severity_encoded (numeric)
# - K-means (Part 2) will use cough_severity_encoded
# - XGBoost (Part 3) will also use cough_severity_encoded (requires numeric input)
covid_ml <- covid_imputed %>%
  select(
    # Target
    covid19_test_results,
    # Numeric features
    temperature,
    # Binary features
    high_risk_exposure_occupation,
    high_risk_interactions,
    labored_respiration,
    rhonchi,
    cough,
    fever,
    sob,
    diarrhea,
    fatigue,
    headache,
    loss_of_smell,
    loss_of_taste,
    runny_nose,
    muscle_sore,
    sore_throat,
    wheezes,
    # Categorical (original, kept for reference)
    cough_severity,
    # Encoded categorical (for K-means and XGBoost)
    cough_severity_encoded
  )

# Final check
cat("Final dataset dimensions:", dim(covid_ml), "\n",
    "Missing values:", sum(is.na(covid_ml)), "\n")
```

```{r save-cleaned-data}
# Save the cleaned dataset for use in subsequent parts
# This file (covid_ml_clean.csv) will be used in:
# - Part 2 (K-means clustering)
# - Part 3 (XGBoost classification)
write.csv(covid_ml, "data/covid_ml_clean.csv", row.names = FALSE)
cat("Cleaned dataset saved to 'data/covid_ml_clean.csv'\n")
cat("This file will be used in Part 2 (K-means) and Part 3 (XGBoost)\n")
```

---

# Summary

In this part, we learned:

1. ✅ **Explore your data** before applying ML algorithms
2. ✅ **Handle missing values** - remove or impute
3. ✅ **Encode categorical variables** - convert to numeric
4. ✅ **Scale features** when using distance-based algorithms (like K-means)

## What's Next?

In the following parts, we'll use this cleaned dataset to:

- **Part 2:** Apply K-means for clustering
- **Part 3:** Apply XGBoost for classification

---

# Session Info

```{r session-info}
sessionInfo()
```