---
title: "Machine Learning Using R - Part 3"
subtitle: "XGBoost"
author: "Gladstone Bioinformatics Core"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction to XGBoost

XGBoost (E**x**treme **G**radient **Boost**ing) is an **ensemble method** that builds decision trees (models that make predictions using if-then rules) **sequentially**, where each new tree tries to correct the errors made by the trees before it.

Two ideas make this work:  
1. **Gradient boosting**: Each new tree is fit to the residual errors of the current ensemble, gradually improving predictions  
2. **Regularization**: Penalty terms (on tree complexity and leaf weights) keep individual trees simple and help prevent overfitting  

This sequential, error-correcting process often makes XGBoost more accurate than bagging-based methods like Random Forest; though it has more hyperparameters to tune!

---

# Load Libraries and Data

```{r libraries}
# Core libraries
library(tidyverse)
library(caret)
library(xgboost)
library(ggplot2)

# Set seed for reproducibility
set.seed(321)
```

```{r load-data}
# Load the cleaned dataset from Part 1
covid_ml <- read.csv("data/covid_ml_clean.csv")

# Convert target to factor
covid_ml$covid19_test_results <- as.factor(covid_ml$covid19_test_results)

# Remove cough_severity - XGBoost requires numeric input
# We'll use the ordinal-encoded cough_severity_encoded instead
covid_ml <- covid_ml %>% select(-cough_severity)
```

**Note:** Unlike Random Forest, XGBoost requires **numeric input** for all predictors - it can't split on a raw categorical/factor variable directly. So instead of the categorical `cough_severity`, we use the ordinal-encoded `cough_severity_encoded` (No cough=0 < Mild=1 < Moderate=2 < Severe=3) that was already prepared in Part 1.

---

# Prepare Data

## Check Class Distribution

First, let's examine the class distribution to identify any imbalance:

```{r check-class-distribution}
# Check class distribution
class_dist <- table(covid_ml$covid19_test_results)
print(class_dist)

cat("\nClass proportions:\n")
prop.table(class_dist)
```

**Class Imbalance Problem:**
With only 68 positives out of 3045 rows (~2.2%), the model might learn to always predict "Negative" and still achieve high accuracy. This is problematic because we care about correctly identifying positive cases!

## Balanced Train-Test Split

To handle class imbalance, we'll use **balanced sampling** for training:  
- **Training set**: Randomly select 50 positive and 50 negative cases (balanced)  
- **Test set**: Use all remaining data (maintains original class distribution)  

This approach helps the model learn from both classes equally during training.

```{r train-test-split}
# Separate data by class
positive_indices <- which(covid_ml$covid19_test_results == "Positive")
negative_indices <- which(covid_ml$covid19_test_results == "Negative")

cat("Total positive cases:", length(positive_indices), "\n")
cat("Total negative cases:", length(negative_indices), "\n")

# Randomly sample 50 from each class for training
set.seed(123)  # For reproducibility
train_positive_indices <- sample(positive_indices, min(50, length(positive_indices)))
train_negative_indices <- sample(negative_indices, 50)

# Combine training indices
train_indices <- c(train_positive_indices, train_negative_indices)

# Create balanced training set
train_data <- covid_ml[train_indices, ]

# Create test set with remaining data
test_data <- covid_ml[-train_indices, ]

# Shuffle training data (optional, but good practice)
train_data <- train_data[sample(nrow(train_data)), ]

cat("\nTraining set:")
print(table(train_data$covid19_test_results))

cat("\nTest set:")
print(table(test_data$covid19_test_results))

cat("\nTraining set size:", nrow(train_data), "\n",
    "Test set size:", nrow(test_data), "\n")
```

**Why Balanced Training?**  
- Prevents the model from being biased toward the majority class  
- Forces the model to learn patterns from both classes  
- Test set maintains original distribution, giving realistic performance estimates  

**Note:** We don't need to scale features for XGBoost! Tree-based methods are invariant to monotonic transformations of features and only the *order* of values matters for choosing a split, not their scale.

## Convert to XGBoost's Matrix Format

The `xgboost` package works directly with numeric matrices (via its `xgb.DMatrix` format) rather than data frames, and expects a numeric 0/1 label instead of a factor:

```{r to-dmatrix}
# Separate predictors (as a numeric matrix) from the target (as 0/1)
train_x <- as.matrix(train_data %>% select(-covid19_test_results))
train_y <- ifelse(train_data$covid19_test_results == "Positive", 1, 0)

test_x <- as.matrix(test_data %>% select(-covid19_test_results))
test_y <- ifelse(test_data$covid19_test_results == "Positive", 1, 0)

# Wrap as xgb.DMatrix - xgboost's optimized internal data structure
dtrain <- xgb.DMatrix(data = train_x, label = train_y)
dtest <- xgb.DMatrix(data = test_x, label = test_y)
```

---

# Train a Simple XGBoost Model

```{r simple-xgb}
# Train a simple XGBoost model with a single, fixed set of hyperparameters
# (these match xgboost's own defaults) - no cross-validation yet, just a
# quick first look at the model before we tune it
xgb_simple <- xgb.train(
  params = list(
    objective = "binary:logistic",  # Binary classification
    max_depth = 6,                  # Max depth of each tree
    eta = 0.3,                      # Learning rate
    eval_metric = "logloss"
  ),
  data = dtrain,
  nrounds = 100,                    # Number of boosting rounds (trees)
  verbose = 0
)

# View the model summary
print(xgb_simple)
```

**No Out-of-Bag Error Here:** Random Forest gets a "free" validation estimate from out-of-bag samples because each tree only sees a bootstrap subset of the data. XGBoost's trees are built sequentially on the *full* training set (with reweighting), so there's no OOB analog - instead we rely on cross-validation to estimate performance, which we'll do in the tuning section below.

---

# Variable Importance

XGBoost also provides built-in variable importance, based on **Gain** - how much each feature contributes to reducing the loss across all boosting rounds.

```{r var-importance}
# Get variable importance from our simple XGBoost model
importance_df <- xgb.importance(model = xgb_simple)

# View importance metrics
head(importance_df)
```

```{r importance-plot}
# Plot variable importance
importance_df %>%
  arrange(desc(Gain)) %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(fill = "#3498db") +
  coord_flip() +
  labs(title = "Variable Importance - XGBoost",
       x = "Feature",
       y = "Gain") +
  theme_minimal()
```

---

# Make Predictions

```{r predict-proba}
# Get prediction probabilities first
xgb_proba <- predict(xgb_simple, dtest)
head(xgb_proba)
```

## Turn Probabilities into Predictions

Notice that the model doesn't say "Positive" or "Negative". For each patient it outputs a **probability** of being positive (the numbers above). To get a final prediction, we pick a **threshold**: patients above it are called "Positive", the rest "Negative".

The default threshold is 0.5, but it's often too low here. We trained on a balanced sample (50% positive), so the model expects positives to be common. In the test set only ~2% are positive, so 0.5 tends to flag too many patients (false positives). We'll use a **higher threshold** (0.7) to be more conservative about positive predictions:

```{r custom-threshold}
# Higher threshold = more specific (fewer false positives, but may miss some positives)
# Lower threshold = more sensitive (catch more positives, but more false alarms)

# For imbalanced data, use a higher threshold to reduce false positives
threshold <- 0.7  # Classify as Positive only if probability > 0.7

xgb_predictions <- ifelse(xgb_proba > threshold, "Positive", "Negative")
xgb_predictions <- factor(xgb_predictions, levels = c("Negative", "Positive"))

# Compare different thresholds
xgb_predictions_default <- ifelse(xgb_proba > 0.5, "Positive", "Negative")
xgb_predictions_default <- factor(xgb_predictions_default, levels = c("Negative", "Positive"))

cat("Default (0.5) - Positive predictions:", sum(xgb_predictions_default == "Positive"), "\n",
    "Conservative (0.7) - Positive predictions:", sum(xgb_predictions == "Positive"), "\n")
```

**Why Higher Threshold?**
With only 18 actual positives in the test set, we want to be very confident before predicting "Positive". A threshold of 0.7 means we only predict positive when the model is 70%+ confident, reducing false positives.

---

# Evaluate Model Performance

## Confusion Matrix

A confusion matrix shows how well the model classified each class:

```
                    Predicted
                 Negative  Positive
Actual Negative    TN        FP
       Positive    FN        TP

TN = True Negative  (correctly predicted negative)
TP = True Positive  (correctly predicted positive)
FN = False Negative (missed positive case)
FP = False Positive (false alarm)
```

```{r confusion-matrix}
# Create confusion matrix
cm_xgb <- confusionMatrix(xgb_predictions, test_data$covid19_test_results,
                          positive = "Positive")
print(cm_xgb)
```

## Understanding the Metrics

```{r metrics-explanation}
# Extract key metrics
accuracy <- cm_xgb$overall["Accuracy"]
sensitivity <- cm_xgb$byClass["Sensitivity"]
specificity <- cm_xgb$byClass["Specificity"]
precision <- cm_xgb$byClass["Precision"]  # Positive Predictive Value

cat("KEY METRICS:\n",
    "============\n",
    "Accuracy:", round(accuracy * 100, 2), "%\n",
    "  - Overall correct predictions\n\n",
    "Sensitivity (Recall):", round(sensitivity * 100, 2), "%\n",
    "  - Of all actual positives, how many did we catch?\n",
    "  - Important when missing a positive case is costly\n\n",
    "Specificity:", round(specificity * 100, 2), "%\n",
    "  - Of all actual negatives, how many did we correctly identify?\n\n",
    "Precision (Positive Predictive Value):", round(precision * 100, 2), "%\n",
    "  - Of all predicted positives, how many were actually positive?\n",
    "  - Important when false positives are costly\n")
```

## Visualize Confusion Matrix

```{r cm-plot}
# Convert confusion matrix to data frame for plotting
cm_df <- as.data.frame(cm_xgb$table)

ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), size = 8, color = "white") +
  scale_fill_gradient(low = "#3498db", high = "#9b59b6") +
  labs(title = "Confusion Matrix - XGBoost",
       x = "Actual",
       y = "Predicted") +
  theme_minimal() +
  theme(axis.text = element_text(size = 12))
```

---

# Tune Hyperparameters

## Key Hyperparameters

XGBoost has more hyperparameters than Random Forest. We'll tune the two that matter most for a first pass, and hold the rest fixed at sensible defaults:

- **max_depth**: Maximum depth of each tree
  - Deeper trees can capture more complex patterns, but risk overfitting
  - **We'll tune this** - testing a few depths
- **eta** (learning rate): How much each tree corrects the previous ensemble's errors
  - Smaller values learn more slowly but can generalize better
  - **We'll tune this too**
- **nrounds**: Number of boosting rounds (trees) - we'll keep this fixed at 100

We'll use `xgb.cv()` - xgboost's built-in cross-validation function - to score each combination of `max_depth` and `eta`, using the **same 10 folds** for every combination so the comparison is fair.

```{r tune-xgb}
# Build 10 stratified folds once, shared across every hyperparameter combination
set.seed(456)
cv_folds <- createFolds(train_data$covid19_test_results, k = 10, returnTrain = FALSE)

# Grid of max_depth / eta values to try
tune_grid <- expand.grid(max_depth = c(2, 4, 6), eta = c(0.05, 0.1, 0.3))
tune_grid$mean_auc <- NA_real_

for (i in seq_len(nrow(tune_grid))) {
  cv_result <- xgb.cv(
    params = list(
      objective = "binary:logistic",
      max_depth = tune_grid$max_depth[i],
      eta = tune_grid$eta[i],
      eval_metric = "auc"
    ),
    data = dtrain,
    nrounds = 100,
    folds = cv_folds,
    verbose = 0
  )
  # Best CV AUC reached across boosting rounds for this combination
  tune_grid$mean_auc[i] <- max(cv_result$evaluation_log$test_auc_mean)
}

# View results, best first
xgb_cv_results <- tune_grid %>% arrange(desc(mean_auc))
xgb_cv_results
```

```{r plot-tune}
# Plot max_depth/eta vs performance
ggplot(tune_grid, aes(x = max_depth, y = mean_auc, color = factor(eta))) +
  geom_line() +
  geom_point(size = 3) +
  labs(title = "XGBoost: Tuning max_depth and eta",
       x = "max_depth",
       y = "Mean CV AUC",
       color = "eta") +
  theme_minimal()
```

```{r best-params}
best_params <- xgb_cv_results[1, ]
cat("Best max_depth:", best_params$max_depth, "\n")
cat("Best eta:", best_params$eta, "\n")
cat("Best mean CV AUC:", round(best_params$mean_auc, 3), "\n")
```

---

# Summary

## What We Learned

1. ✅ **XGBoost** builds decision trees sequentially, each one correcting the errors of the ensemble so far
2. ✅ **Gradient boosting + regularization** balances fitting the data well against overfitting
3. ✅ **Numeric input required** - unlike Random Forest, categorical variables need to be encoded first
4. ✅ **No scaling required** - trees are still scale-invariant, even though boosting is sequential
5. ✅ **Variable importance** (via `xgb.importance()`) helps understand which features matter most
6. ✅ **No OOB error** - boosting uses the full training set each round, so we rely on cross-validation instead
7. ✅ **Cross-validation** (via `xgb.cv()`) helps tune hyperparameters like `max_depth` and `eta`
8. ✅ **Class imbalance handling** - balanced sampling for training prevents bias toward majority class

## Workshop Complete!

You've now learned preprocessing, unsupervised learning (K-means), and supervised learning (XGBoost)!

---

# Session Info
```{r session-info}
sessionInfo()
```
