---
title: "AI Fluency for Biologists"
subtitle: "Part 1"
author: "Natalie Gill"
institute: "Gladstone Bioinformatics Core"
date: "TBD"
date-format: MMMM DD, YYYY
format:
  revealjs:
    theme: [default, ai-theme.scss]
    highlight-style: arrow
    code-line-numbers: false
    template-partials:
      - title-slide.html
---

```{r}
#| label: setup
#| include: false
library(countdown)
```

##

<center>*Press the ? key for tips on navigating these slides*</center>

## Introductions

**Natalie Gill**   
Bioinformatician II  


## Poll 1

**What is your level of experience with AI tools like ChatGPT or Claude?**

1. I've never used them
2. I've tried them a few times
3. I use them regularly
4. I use them daily in my work

## Target Audience

- No prior AI or programming experience required
- Designed for biologists at all career stages
- Laptop and a free Claude account are all you need

## Learning Objectives

1. Identify tasks that are well-suited for AI assistance
2. Write clear, effective prompts to get useful results
3. Evaluate AI outputs for accuracy and appropriateness
4. Recognize limitations and potential biases in AI responses
5. Apply ethical best practices when using AI in research

## Part 1:

1. Introduction to AI tools and the 4 Ds framework
2. Module 1: Delegation
3. Break
4. Module 2: Description


# Introduction

## What is AI Fluency?

The ability to work with AI systems in ways that are **effective**, **efficient**, **ethical**, and **safe**

- Not about becoming an AI expert
- About developing practical skills to use AI as a tool in your research
- Like learning to use a new instrument - you need to know its strengths and limitations

## The 4 Ds Framework

A practical approach to working with AI effectively:

1. **Delegation** - Deciding *what* to ask AI to do
2. **Description** - Crafting *how* you ask (prompting)
3. **Discernment** - Evaluating *whether* the output is correct
4. **Diligence** - Using AI *responsibly* and ethically


## {.center}

![](assets/four_ds_diagram.svg){fig-alt="The 4 Ds of AI Fluency: Delegation, Description, Discernment, and Diligence arranged in an interconnected diamond with bidirectional arrows between each node" fig-align="center" width=100%}

## AI as Augmentation, Not Replacement

AI tools are **thinking partners**, not replacements for your expertise

- Your domain knowledge is essential - AI doesn't know your system, your data, or your scientific question
- AI is useful for tasks like brainstorming, summarizing, drafting, and troubleshooting
- You remain responsible for the scientific validity of your work

<!-- 

Note: This is probably well covered by the other slides
## Current AI Landscape {.smaller}

**Where AI can help in biology:**

- Literature search and summarization
- Writing and editing manuscripts, grants, and protocols
- Designing experiments and analysis strategies
- Explaining unfamiliar concepts or statistical methods

**Where caution is needed:**

- AI can "hallucinate" - generating plausible but fabricated references, gene names, or statistics
- Outputs may reflect biases in training data
- Sensitive or unpublished data should not be shared with AI tools without considering privacy policies
- AI-generated figures should not be used in publications - they can misrepresent data and many journals prohibit them -->

## Brief Overview of Tools {.small_custom}

Popular tools include **Claude** (Anthropic), **ChatGPT** (OpenAI), and **Gemini** (Google)

- All are **large language models (LLMs)** - trained on vast text data to understand and generate language
- They share the same core limitations: hallucinations, knowledge cutoffs, and training biases
- Relative performance varies by task and changes frequently
- They do not search the internet by default (some have optional web search)



# Module 1: Delegation

## Goal and Task Awareness {.small_custom}

Before opening any AI tool, clearly define what you need:

- **What is the task?** e.g., summarize a paper, draft a methods section, troubleshoot a protocol
- **What does success look like?** Define the output you want before you start
- **Do you have the expertise to evaluate the result?** If you can't judge whether the AI is right, you probably shouldn't delegate it
- Together with **Platform Awareness**, this helps you decide if AI is the right tool for the job

## Platform Awareness {.small_custom}

Understanding what AI tools can and cannot do:

- **Context window** -- how much text AI can process at once. Exceeding it causes the AI to "forget" earlier content
- **Knowledge cutoff** -- AI has no knowledge after its training date. It won't know recent papers
- **Hallucination** -- AI can confidently state plausible but wrong things (fake citations, invented gene names)
- **Context pollution** -- long conversations accumulate errors; start a new chat for new tasks
- **Extended thinking / research modes** -- newer features for complex reasoning or searching external sources


## Choosing an AI Tool

**Before choosing a tool, consider the privacy policy:**

- What data is stored? Is it used to train future models?
- Does your institution have approved tools or data use agreements?
- Never share sensitive, unpublished, or patient-derived data without checking your institution's policies

## {.center}

![](assets/task_delegation_modalities.svg){fig-alt="Three modalities of task delegation shown as cards: Automation (you define, AI executes), Augmentation (you and AI collaborate as thinking partners), and Agency (you configure, AI acts independently), with bar charts showing relative human versus AI involvement" fig-align="center" width=95%}

## Modalities of Task Delegation {.smaller}

- **Automation** -- You define exactly what needs to be done, and the AI executes it

:::{.example-indented}
"Rewrite this protocol section to follow the journal's formatting guidelines, do not change the content - just the formatting."
:::

- **Augmentation** -- You and AI collaborate iteratively as thinking partners

:::{.example-indented}
Drafting and refining an abstract together over several rounds of feedback
:::

- **Agency** -- You configure the AI's role and knowledge, then let it act independently

:::{.example-indented}
Giving AI your experimental goals and reagent constraints, and letting it independently design and compare multiple protocol options
:::

In practice, **Augmentation** is the most productive mode for research tasks

## Exercise: Identify Your AI Tasks {.small_custom}

**Think about your own work and identify 2-3 tasks where AI could help.**

For each task, consider:

1. What is the task? (e.g., literature review, writing, data wrangling)
2. Which mode fits best -- Automation, Augmentation, or Agency?
3. Could you evaluate whether the AI's output is correct?

Enter tasks [here](https://app.sli.do/event/3sDxDAGKp3d6j1LfuKN4Si).


# 10 min break

<center>

```{r}
#| echo: false
countdown::countdown(minutes = 10,
                     seconds = 0,
                     color_border = "black",
                     color_running_background = "#47d193",
                     color_finished_background = "#a3184e",
                     padding = "50px",
                     margin = "5%",
                     font_size = "4em",
       style = "position: relative; width: min-content;")
```

</center>


# Module 2: Description

## 6 Prompting Techniques Overview {.small_custom}

1. **Provide Context** -- Give background information the AI needs
2. **Show Examples** -- Demonstrate the format or style you want
3. **Specify Output Constraints** -- Define length, format, audience, and structure
4. **Break Complex Tasks into Steps** -- Provide sequential instructions
5. **Ask It to Think First** -- Use extended thinking modes
6. **Define the AI's Role** -- Set perspective

These techniques map to the Description competency: Product, Process, and Performance Description

## 1) Provide Context {.small_custom}

Give the AI the background information it needs to do a good job.

- What field or subfield are you working in?
- What is the purpose of this task?
- What has already been done?


:::{.bad-example}
"Summarize this paper"
:::

:::{.good-example}
"I'm a graduate student in cardiac biology preparing for a journal club. Summarize this paper's main findings, focusing on the signaling pathways involved in cardiomyocyte differentiation."
:::

## 2) Show Examples

Show the AI what you want by providing one or more examples of the desired output.

- Useful for defining **format**, **style**, or **level of detail**
- Also called "few-shot prompting" -- the AI learns from the pattern you demonstrate


:::{.example}
"Summarize each paper in this format:

[Author et al., Year] -- [One-sentence finding]. Method: [Key technique]. Relevance: [Why it matters to my project]."
:::

## 3) Specify Output Constraints

Tell the AI exactly what the output should look like:

- **Length** -- "in 200 words," "in 3 bullet points"
- **Format** -- "as a table," "as a numbered list," "in paragraph form"
- **Audience** -- "for a PI," "for a first-year graduate student," "for a grant review panel"
- **Structure** -- "organize by: (1) hypothesis, (2) methods, (3) key results"

The more specific your constraints, the closer the output will match what you need

## Bad Prompt vs. Good Prompt {.small_custom}

:::{.bad-example}
"Summarize this paper"
:::

:::{.good-example}
"Summarize this cell biology paper's experimental methodology in exactly 200 words, focusing only on protein-protein interaction techniques (e.g., co-IP, pull-downs, FRET, crosslinking). Organize your summary by:  

1. Primary techniques used 
2. Key reagents
3. Validation methods

Write for an audience of post-docs and graduate students."
:::

The good prompt uses **context**, **output constraints**, and **structure** to guide the AI

## 4) Break Complex Tasks into Steps {.small}

For complex tasks, give the AI a step-by-step process to follow.


:::{.example}
"Analyze this paper in the following steps:

Step 1: Identify the main hypothesis

Step 2: List the experimental methods used

Step 3: Summarize the key findings

Step 4: Evaluate whether the conclusions are supported by the data"
:::

This is called **Process Description** -- defining *how* the AI should approach the task

## 5) Let It Think Deeply {.small}

Most AI tools now have a **thinking** or **reasoning** mode — use it for complex or multi-part questions

- **Don't** need to say "think step by step"
- **Do** toggle on extended thinking for harder questions; skip it for simple tasks

:::{.example}
*[Turn on extended thinking / reasoning mode]*

"Here is a paper on CRISPR-based gene therapy for sickle cell disease. What are the key methodological limitations, and how might they affect the authors' conclusions?"
:::

## 6) Define the AI's Role {.small}

Set the AI's perspective to shape its responses.


:::{.example}
- "Explain this as a journal club presenter would to a mixed audience of biologists"
- "Write as if you are teaching a graduate student new to this subfield"
- "Act as a critical reviewer evaluating this manuscript for a high-impact journal"
:::

This is **Performance Description** -- defining how the AI should behave during your interaction (concise vs. detailed, supportive vs. critical)

## Bonus: Ask AI to Help with Your Prompt {.small}

You can ask the AI to help you write a better prompt -- this is called **meta-prompting**.


:::{.example}
"I'm trying to analyze this paper's methodology for my literature review. Can you help me craft a better prompt to get a useful summary?"
:::

This is especially useful when:

- You're not sure how to structure your request
- You want to explore what's possible before committing to a specific approach
- Your initial prompt isn't giving you good results

## Exercise: Prompting {.small_custom}

**Use the task you identified in the Delegation exercise and write a prompt using at least 3 of the 6 techniques.**

**Steps:**

1. **Draft your prompt** using at least 3 techniques (3 min)
2. **Run it in Claude** and review the output (3 min)
3. **Iterate** -- refine your prompt and try again if needed (2 min)
4. **Reflect:** Which techniques helped most? What would you change? (2 min)


*Tip: Start with context + constraints + role, then add steps or examples if needed.*


# End of Part 1


## Schedule for Part 2

1. Module 3: Discernment - Evaluating AI outputs
2. Module 4: Diligence - Responsible AI use
3. Practical application exercise
4. Q&A and discussion
5. Wrap-up and resources

## Workshop survey
- Please fill out our workshop survey so we can continue to improve these workshops

## Upcoming Workshops {.small}

*Details coming soon -- check our website for the latest schedule*

## Acknowledgments

This workshop adapts the AI Fluency framework developed by Rick Dakan, Joseph Feller, and Anthropic (2025).

Released under the [CC BY-NC-SA 4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/) license.
