---
title: "AI Fluency for Biologists"
subtitle: "Part 1"
author: "Natalie Gill"
institute: "Gladstone Bioinformatics Core"
date: "TBD"
date-format: MMMM DD, YYYY
format:
  revealjs:
    theme: [default, ai-theme.scss]
    highlight-style: arrow
    code-line-numbers: false
    template-partials:
      - title-slide.html
---

```{r}
#| label: setup
#| include: false
library(countdown)
```

##

<center>*Press the ? key for tips on navigating these slides*</center>

## Introductions

**Natalie Gill**   
Bioinformatician II  


## Poll 1

**What is your level of experience with AI tools like ChatGPT or Claude?**

1. I've never used them
2. I've tried them a few times
3. I use them regularly
4. I use them daily in my work

## Target Audience

- No prior AI or programming experience required
- Designed for biologists at all career stages
- Laptop and a free Claude account are all you need

## Learning Objectives

1. Identify tasks that are well-suited for AI assistance
2. Write clear, effective prompts to get useful results
3. Evaluate AI outputs for accuracy and appropriateness
4. Recognize limitations and potential biases in AI responses
5. Apply ethical best practices when using AI in research

## Part 1:

1. Introduction to AI tools and the 4 Ds framework
2. Module 1: Delegation
3. Break
4. Module 2: Description


# Introduction

## What is AI Fluency?

The ability to work with AI systems in ways that are **effective**, **efficient**, **ethical**, and **safe**

- Not about becoming an AI expert
- About developing practical skills to use AI as a tool in your research
- Like learning to use a new instrument - you need to know its strengths and limitations

## The 4 Ds Framework

A practical approach to working with AI effectively:

1. **Delegation** - Deciding *what* to ask AI to do
2. **Description** - Crafting *how* you ask (prompting)
3. **Discernment** - Evaluating *whether* the output is correct
4. **Diligence** - Using AI *responsibly* and ethically

## AI as Augmentation, Not Replacement

AI tools are **thinking partners**, not replacements for your expertise

- Your domain knowledge is essential - AI doesn't know your system, your data, or your scientific question
- AI is useful for tasks like brainstorming, summarizing, drafting, and troubleshooting
- You remain responsible for the scientific validity of your work


## Current AI Landscape {.small}

**Where AI can help in biology:**

- Literature search and summarization
- Writing and editing manuscripts, grants, and protocols
- Designing experiments and analysis strategies
- Explaining unfamiliar concepts or statistical methods

**Where caution is needed:**

- AI can "hallucinate" - generating plausible but fabricated references, gene names, or statistics
- Outputs may reflect biases in training data
- Sensitive or unpublished data should not be shared with AI tools without considering privacy policies
- AI-generated figures should not be used in publications - they can misrepresent data and many journals prohibit them

## Brief Overview of Tools

Popular tools include **Claude** (Anthropic), **ChatGPT** (OpenAI), and **Gemini** (Google)

- All are **large language models (LLMs)** - trained on vast text data to understand and generate language
- They share the same core limitations: hallucinations, knowledge cutoffs, and training biases
- Relative performance varies by task and changes frequently - no single tool is consistently "best"
- They do not search the internet by default (some have optional web search)


## Choosing an AI Tool

**Before choosing a tool, consider the privacy policy:**

- What data is stored? Is it used to train future models?
- Does your institution have approved tools or data use agreements?
- Never share sensitive, unpublished, or patient-derived data without checking your institution's policies


# Module 1: Delegation

## Goal and Task Awareness

*TODO: Define your task before opening AI*

## Platform Awareness

*TODO: Context windows, knowledge cutoffs, hallucination risks*

## Three Modalities of Task Delegation

*TODO: Automation, Augmentation, Agency*

## Exercise: Identify Your AI Tasks

*TODO: Identify 2-3 tasks from own work suitable for AI*


# 10 min break

<center>

```{r}
#| echo: false
countdown::countdown(minutes = 10,
                     seconds = 0,
                     color_border = "black",
                     color_running_background = "#47d193",
                     color_finished_background = "#a3184e",
                     padding = "50px",
                     margin = "5%",
                     font_size = "4em",
       style = "position: relative; width: min-content;")
```

</center>


# Module 2: Description

## 6 Prompting Techniques Overview

*TODO: List the 6 techniques*

## Technique 1: Provide Context

*TODO: Background information the AI needs*

## Technique 2: Show Examples

*TODO: Demonstrate format or style*

## Technique 3: Specify Output Constraints

*TODO: Length, format, audience, structure*

## Bad Prompt vs. Good Prompt

*TODO: Side-by-side comparison*

## Technique 4: Break Complex Tasks into Steps

*TODO: Sequential instructions*

## Technique 5: Ask It to Think First

*TODO: Reason before answering*

## Technique 6: Define the AI's Role

*TODO: Set expertise and perspective*

## Bonus: Ask AI to Help with Your Prompt

*TODO: Meta-prompting demo*

## Exercise: Craft a Prompt

*TODO: Apply at least 3 techniques, run through Claude, evaluate*


# End of Part 1


## Schedule for Part 2

1. Module 3: Discernment - Evaluating AI outputs
2. Module 4: Diligence - Responsible AI use
3. Practical application exercise
4. Q&A and discussion
5. Wrap-up and resources

## Workshop survey
- Please fill out our workshop survey so we can continue to improve these workshops

## Upcoming Workshops {.small}

*TODO: Add upcoming workshop details*

## Acknowledgments

This workshop adapts the AI Fluency framework developed by Rick Dakan, Joseph Feller, and Anthropic (2025).

Released under the [CC BY-NC-SA 4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/) license.
