---
title: "AI Fluency for Biologists"
subtitle: "Part 2"
author: "Natalie Gill"
institute: "Gladstone Bioinformatics Core"
date: "TBD"
date-format: MMMM DD, YYYY
format:
  revealjs:
    theme: [default, ai-theme.scss]
    highlight-style: arrow
    code-line-numbers: false
    template-partials:
      - title-slide.html
---

```{r}
#| label: setup
#| include: false
library(countdown)
```

##

<center>*Press the ? key for tips on navigating these slides*</center>

# Introductions
**Natalie Gill**
Bioinformatician II


# Schedule

1. Recap of Part 1
2. Module 3: Discernment — Evaluating AI outputs
3. Module 4: Diligence — Responsible AI use
4. Practical application exercise
5. Q&A and discussion
6. Wrap-up and resources


## Recap: Part 1

*TODO: Quick recap of Delegation and Description modules*


# Module 3: Discernment

## Three Types of Discernment

*TODO: Product, Process, Performance discernment*

## Product Discernment

*TODO: Accuracy, Precision, Completeness, Appropriateness*

## Common AI Errors in Science

*TODO: Hallucinated citations, oversimplified mechanisms, etc.*

## Exercise: Spot the Errors

*TODO: Review pre-generated AI summary, identify 3-5 issues*

## Process Discernment

*TODO: Red flags and green flags*

## When to Stop Using AI for a Task

*TODO: Knowing when to stop*

## Performance Discernment: Verification Workflow

*TODO: 5-step verification process*

## Demo: Fact-Checking an AI Summary

*TODO: Live demonstration*


# 10 min break

<center>

```{r}
#| echo: false
countdown::countdown(minutes = 10,
                     seconds = 0,
                     color_border = "black",
                     color_running_background = "#47d193",
                     color_finished_background = "#a3184e",
                     padding = "50px",
                     margin = "5%",
                     font_size = "4em",
       style = "position: relative; width: min-content;")
```

</center>


# Module 4: Diligence

## Creation Diligence

*TODO: Copyright, plagiarism, scientific integrity, data privacy*

## Transparency Diligence

*TODO: When and how to acknowledge AI use*

## Deployment Diligence

*TODO: Quality control process*

## Journal AI Policies

*TODO: Common journal policy themes*


# Practical Application

## Hands-On Exercise

*TODO: Create 150-200 word paper summary using techniques*

## Exercise Steps

*TODO: Draft prompt, run it, evaluate, reflect*

## Share Your Results

*TODO: 2-3 volunteers share prompts and results*


# Q&A and Discussion

## Open Discussion

*TODO: Questions, troubleshooting, edge cases*


# Wrap-Up

## Key Takeaways

*TODO: The 4 Ds summary*

## Resources

*TODO: Links and references*


# End of Part 2

## Workshop survey
- Please fill out our workshop survey so we can continue to improve these workshops

## Upcoming Workshops {.small}

*TODO: Add upcoming workshop details*
