---
title: "Intro to CoreHPC"
subtitle: "Part 1"
author: "Natalie Gill"
institute: "Gladstone Bioinformatics Core"
date: "TBD"
date-format: MMMM DD, YYYY
format:
  revealjs:
    theme: corehpc-theme.scss
    highlight: default
    code-line-numbers: false
    template-partials:
      - title-slide.html
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(comment = "")
```

##

<center>*Press the ? key for tips on navigating these slides*</center>

## Introductions


&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;**Natalie Gill**
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;*Bioinformatician II*

## Target Audience

-   Prior experience with UNIX command-line
-   No prior experience on computing clusters

## Part 1:

1.  What is CoreHPC?
2.  Accessing CoreHPC
3.  Storage
4.  Data Transfer
5.  Software with OpenHPC Modules
6.  Python Environments with Miniforge3
7.  Containers


# What is CoreHPC?

## Overview {.small-bullets}

-   An institutional HPC cluster for **UCSF researchers**
-   Built and maintained by **UCSF IT Academic Research Services**, in partnership with **Gladstone IT**
-   Originally focused on **AI / Data Science**, now expanding to general-purpose HPC
-   Set to **replace Wynton** (Wynton is scheduled to sunset by Feb 2027)

## Architecture

![](slide_materials/CoreHPC_diagram.svg){fig-alt="CoreHPC architecture diagram showing bastion nodes, login node, compute nodes, and Globus endpoints" .nostretch fig-align="center" width=80%}

<center>*Connect through a **bastion**, submit jobs from the **login node**, and SLURM runs them on **compute nodes**. **Globus** moves data in and out.*</center>

## A Two-Tenant System {.small-bullets}

-   CoreHPC is **one HPC platform** serving **two separate tenants**: **UCSF campus** and **Gladstone**
-   Each tenant has its **own** bastion, login nodes, **storage**, and **compute nodes**
-   Both run the **same setup** (Linux, OpenHPC modules, Apptainer, SLURM), so this workshop applies to either
-   This is why some hostnames and paths differ by institution: **use the ones for your tenant**

## Partitions {.small-bullets}

-   **`cpu`** : high-core-count CPU nodes for general compute
-   **`small_gpu`** : single-GPU jobs (NVIDIA L40s / H100NVL)
-   **`large_gpu`** : large single-node GPU jobs (NVIDIA H200)
-   **`pod`** : multi-GPU SuperPod (8x H200 with InfiniBand), **by request only**
-   **`community cpu`** : high-throughput CPU, no InfiniBand (Wynton-like)

*Partition names and availability change as the cluster grows. Confirm with `sinfo`.*

## Hardware Highlights {.small-bullets}

-   **CPU nodes:** dual AMD EPYC (84+ cores), 1.5 TB RAM
-   **small_gpu:** dual **L40s (48 GB)** or **H100NVL**, 384 GB RAM, 32 cores
-   **large_gpu:** **H200** GPUs on high-memory nodes
-   **pod:** 8x **H200 SXM** per node, 2 TB RAM, InfiniBand across GPUs
-   Every compute node also has fast **local NVMe scratch** (2 to 12 TB)

*Specs as of mid-2026; see the current CoreHPC docs for the latest.*


# Accessing CoreHPC

## Prerequisites {.small-bullets}

-   A **UCSF campus account** 
-   A **CoreHPC-provisioned account** (requested via your PI / lab)
-   On the right network:
    -   **UCSF:** UCSF network or VPN
    -   **Gladstone:** Gladstone network (wired, Gladstone-MB, or Ivanti VPN)
-   **MFA / DUO** enrolled on your phone

## Step 1: SSH to a Bastion Node {.smaller}

*A **bastion** is a hardened gateway: it authenticates you and lets you SSH onward. It has no storage or compute of its own.*

:::: {.columns}

::: {.column width="48%"}
**UCSF**

```bash
ssh <user>@chpc-ucsf-bastion-vm1.corehpc.ucsf.edu
# alternate: chpc-ucsf-bastion-vm2.corehpc.ucsf.edu
```
:::

::: {.column width="48%"}
**Gladstone**

```bash
ssh <user>@chpc-gs-bastion-vm1.corehpc.ucsf.edu
```
:::

::::

-   Enter your **UCSF MyAccess password**, then approve the **DUO** push

## Step 2: Hop to the Login Node {.smaller}

From the bastion, SSH to the login node:

:::: {.columns}

::: {.column width="48%"}
**UCSF**

```bash
ssh chpc-ucsf-login-vm1
```
:::

::: {.column width="48%"}
**Gladstone**

```bash
ssh chpc-gs-login-vm1
```
:::

::::

```{r, engine='bash', eval=TRUE, results='markup', comment=NA, highlight=TRUE, echo=FALSE}
echo '[alice@laptop ~]$ ssh alice@chpc-gs-bastion-vm1.corehpc.ucsf.edu
Password:
Duo two-factor login for alice
Passcode or option (1-1): 1
Success. Logging you in...
[alice@chpc-gs-bastion-vm1 ~]$ ssh chpc-gs-login-vm1
[alice@chpc-gs-login-vm1 ~]$'
```

## The Login Node {.small-bullets}

-   Your launch pad: **submit jobs** and **stage data**, not for heavy compute
-   The **only** node with internet (via a proxy): GitHub, PyPI, conda, Hugging Face, Apptainer registries
-   Do **not** run analyses here. Use `salloc` (interactive) or `sbatch` (batch) instead
-   Shares your home directory and scratch with the compute nodes

## The Bastion Nodes {.small-bullets}

-   A **gateway only**: authentication and SSH forwarding, no storage or compute
-   `scp` / `rsync` through the bastion are **blocked** by security policy
-   Move data with **Globus** or via a mounted **FAC / HIVE share** instead


# Storage

## Storage Layout {.smaller}

| Location | Path | Size / policy |
|----------|------|---------------|
| Home | `/home/remote/$USER` (shared) | 20 GB |
| Scratch | `/scratch/user/$USER` *(Gladstone: `/mnt/scratch/...`)* | 50 TB cap, purged after 30 days |
| Group scratch | `/scratch/group/<group>` | shared, volatile |
| Persistent | FAC capacity storage *(Gladstone: HIVE `/mnt/gladstone/<pi>/<share>`)* | persistent, backed up |
| Local | per-node NVMe scratch | fast, lasts only for the job |

## Home Directory {.small-bullets}

-   **20 GB** quota, shared across login and compute nodes (not the bastion)
-   Good for scripts, configs, small files, and `environment.yml` recipes
-   Too small for conda envs, data, or container images (push those to scratch / HIVE)

## Scratch {.small-bullets}

-   Large, fast shared space for **active job data**
-   Path: `/scratch/user/$USER` *(Gladstone: `/mnt/scratch/user/$USER`)*
-   **50 TB** per-user cap; **anything older than 30 days is automatically deleted**
-   Treat it as **volatile**: never your only copy

## FAC Capacity Storage {.small-bullets}

-   Persistent project storage for data you keep beyond 30 days
-   Request a share: <https://tiny.ucsf.edu/getfac>
-   **Gladstone:** your **HIVE** lab share (`/mnt/gladstone/<pi>/<share>`) is already mounted and **backed up**, use it for persistence

## Storage Advice {.small-bullets}

-   Scratch is volatile: keep `environment.yml`, scripts, and small inputs on home / HIVE
-   Use **FAC / HIVE** for anything you cannot afford to lose
-   Avoid huge directories of tiny files on shared storage (it slows the filesystem for everyone). Prefer a single container image (see the [container guide](Containers_on_CoreHPC.html))


# Data Transfer

## Transfer Options {.small-bullets}

-   **Globus** (recommended) for moving data in and out of CoreHPC
-   **`cp` / `rsync` from the login node** if your data is on a mounted FAC / HIVE share
-   `scp` / `rsync` **through the bastion are blocked**

## Globus {.small-bullets}

-   Log in at <https://globus.org> with your **UCSF CILogon** identity
-   Endpoints: **"UCSF CoreHPC - User Home"** and **"UCSF CoreHPC - User Scratch"**
-   For your laptop, install **Globus Connect Personal** (use your UCSF ORG / CILogon identity)
-   Best for large files (> 1 GB) and cross-institution transfers

## FAC Share Copy

If your data is on a mounted share, just copy it on the login node:

```bash
cp -R /mnt/fac/<share>/<file> /mnt/scratch/group/<share>/
# or
rsync -av /mnt/fac/<share>/<file> /mnt/scratch/group/<share>/
```

-   **Gladstone:** data already on **HIVE** (`/mnt/gladstone/...`) needs no move, it is mounted on CoreHPC

## Hands-on

-   Transfer a small file into your scratch space using Globus
-   We will walk through a Globus transfer together


# Software with OpenHPC Modules

## Module Basics {.small-bullets}

-   Cluster software is provided as **OpenHPC modules**
-   A module sets up your `PATH` and environment for a given tool and version
-   Available consistently on the **login and compute nodes**
-   Wynton's **CBI modules are ported over** (e.g. `module load CBI r`)

## Common Commands

```bash
module avail              # list available modules
module load miniforge3    # load a module
module list               # show loaded modules
module unload miniforge3  # unload it
module --help
```

## Example Available Modules {.small-bullets}

```{r, engine='bash', eval=TRUE, results='markup', comment=NA, highlight=TRUE, echo=FALSE}
echo '[alice@chpc-gs-login-vm1 ~]$ module avail

  apptainer/1.4.1      gnu13/13.2.0          nvidia/cuda/12.9.1
  cmake/4.0.0          hwloc/2.12.0          nvidia/cudnn/9.10_cuda_12.9
  cuda/25.1            libfabric/1.18.0      nvidia/nvhpc/25.5
  autotools            miniforge3/25.11.0    ucx/1.18.0
  papi/6.0.0           spack/0.23.1          prun/2.2'
```

*Module versions shown are current as of mid-2026.*


# Python Environments with Miniforge3

## What is Miniforge? {.small-bullets}

-   A minimal **conda** distribution that defaults to the **conda-forge** channel
-   Lets you create isolated **environments**, each with its own Python and packages
-   Available as a module: `module load miniforge3`

## Setup (Once) {.smaller}

Home is only 20 GB, so point conda's environments and package cache at scratch:

```bash
module load miniforge3
conda init bash          # then restart your shell
conda config --append envs_dirs /mnt/scratch/user/$USER/envs
conda config --append pkgs_dirs /mnt/scratch/user/$USER/pkgs
```

*UCSF (non-Gladstone): use `/scratch/user/$USER/...` instead.*

## Creating an Environment

```bash
conda create -n project1 python=3.11
conda activate project1
conda install numpy pandas scikit-learn
```

## Exporting an Environment {.small-bullets}

Scratch is volatile, so capture a recipe you can rebuild from:

```bash
conda env export --no-builds > environment.yml   # save to home or HIVE
conda env create -f environment.yml              # rebuild anytime
```

-   For long-lived environments, build an **Apptainer image** and store the `.sif` on HIVE instead. See the [container guide](Containers_on_CoreHPC.html).


# Containers

## Motivation {.small-bullets}

-   Package your code with the **exact versions** of every tool, so it runs the same anywhere
-   **Reproducible**, isolated from the host OS, and great for hard-to-install software
-   One `.sif` file is far friendlier to shared storage than a conda env's **millions of tiny files**

## Apptainer on CoreHPC {.small-bullets}

-   Load the module: `module load apptainer`
-   Pull and run images:

```bash
apptainer pull docker://<user>/<image>:<tag>   # creates image_tag.sif
apptainer exec image.sif <command>
apptainer shell image.sif                       # interactive shell
```

-   **Pull / build on the login node** (the only node with internet)

## Example: Pull a Container

```bash
# On the login node:
apptainer pull docker://natalie23gill/hello-world:1.0
apptainer exec hello-world_1.0.sif hi
```

-   **Building your own image** (from a conda env or Dockerfile): see the [container guide](Containers_on_CoreHPC.html)


# Polls

## Poll 1

Which node should you run heavy compute on?

1.  A bastion node
2.  The login node
3.  A compute node (via `sbatch` / `salloc`)
4.  Any of the above

## Poll 2

Files in your home directory are kept forever, so it is safe to store your only copy there.

1.  True
2.  False


# End of Part 1

## Workshop survey

- If you are not attending part 2, please fill out our [workshop survey](TBD) so we can continue to improve these workshops

## Upcoming Data Science Training Program Workshops {.smaller}

*To be filled in closer to the workshop date.*
