---
title: 'Hands-on component of Single-cell RNA-seq workshop: Sessions 1-2'
author: "Krishna Choudhary"
date: "3/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Introduction

We'll review the current practices for some of the common steps in scRNA-seq data analysis in the Sessions 1-2. We'll discuss different practices for each step and the assumptions underlying various tools and their limitations. This document provides an exposure to one of the popular tools for such analysis. As we'll discuss, the right choice of method in any application depends on a number of factors, including the biological systems under study and the characteristics of the data in hand. For the purpose of our workshop, we'll limit the hands-on component to the Seurat package in R. In general, analysis might require multiple tools in different languages and/or novel development. For more, please see the slide deck in materials. 

The following is based on [this vignette](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html) from the Seurat developers.

## Setup the working environment

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(Seurat)
library(patchwork)
```

## Load the data

Please ensure that the directory named "pbmc3k_data" in the workshop materials is in the same directory as this .Rmd file.

```{r}
data <- CreateSeuratObject(counts = Read10X("pbmc3k_data"),
                           project = "Hello_scWorld",  #Name this whatever.
                           min.cells = 3,  # Don't keep genes observed in fewer than 3 cells
                           min.features = 200)  # Don't keep cells with fewer than 200 genes
```

## Filter poor quality or uninteresting cells

```{r}
#Assess percent of mitochondrial counts in each cell 
data[["percent_mt"]] <- PercentageFeatureSet(object = data, 
                                             pattern = "^MT-")

#Violin plot
VlnPlot(object = data, 
        features = c("nFeature_RNA", 
                   "nCount_RNA", 
                   "percent_mt"), 
        ncol = 3)

#Other plotting otions
plot1 <- FeatureScatter(object = data, 
                        feature1 = "nCount_RNA", 
                        feature2 = "percent_mt")
plot2 <- FeatureScatter(object = data,
                        feature1 = "nCount_RNA", 
                        feature2 = "nFeature_RNA")
plot1 + plot2

#Subset data
data <- subset(x = data, 
               subset =  nFeature_RNA > 200 & 
                 nFeature_RNA < 2500 & 
                 percent_mt < 5)

VlnPlot(object = data, 
        features = c("nFeature_RNA",
                   "nCount_RNA",
                   "percent_mt"),
        ncol = 3)

```

## Normalization

```{r}
data <- NormalizeData(object = data, 
                      normalization.method = "LogNormalize", 
                      scale.factor = 10000)

```

## Feature selection

```{r}
data <- FindVariableFeatures( object = data, 
                              selection.method = "vst", 
                              nfeatures = 2000)

#View the 10 most highly variable genes
top10 <- head(x = VariableFeatures(object = data), 10)
print(top10)

#Seurat allows plotting variable features with and without labels
plot1 <- VariableFeaturePlot(object = data)
plot2 <- LabelPoints(plot = plot1, 
                     points = top10, 
                     repel = TRUE)
plot1 
plot2

```

## Dimensionality reduction

### Linear dimensionality reduction
```{r}
#By default Seurat only scales the variable features.
#Explicit input required to rescale all the genes
scale_genes <- rownames(data)
data <- ScaleData(object = data, 
                  features = scale_genes)

# Use the highly variable genes to find principal components
data <- RunPCA(object = data,
               features = VariableFeatures(object = data),  
               verbose = FALSE)  

#Examine and visualize PCA results a few different ways
print(x = data[["pca"]], dims = 1:5, nfeatures = 5)

DimPlot(data, reduction = "pca")

data <- JackStraw(object = data, num.replicate = 100)
data <- ScoreJackStraw(object = data, dims = 1:20)
JackStrawPlot(data, dims = 1:15)

ElbowPlot(data)

VizDimLoadings(object = data, dims = 1:2, reduction = "pca")

```

### Nonlinear dimensionality reduction
```{r}
data <- RunUMAP(data, dims = 1:10)
data <- RunTSNE(data, dims = 1:10)

DimPlot(data, reduction = "umap")
DimPlot(data, reduction = "tsne")
```

## Clustering
```{r}
data <- FindNeighbors(data, dims = 1:10)
data <- FindClusters(data, resolution = 0.5)

DimPlot(data, reduction = "umap", label = TRUE) 
DimPlot(data, reduction = "tsne", label = TRUE)
DimPlot(data, reduction = "pca", label = TRUE)
```

### Save the Seurat object
```{r}
test <- data[, 1:10 ]
saveRDS(test, file = "hello_scWorld.rds")
```

## Find markers
```{r}
# find all markers of cluster 1
cluster1.markers <- FindMarkers(data, 
                                ident.1 = 1, 
                                min.pct = 0.25)
head(cluster1.markers, n = 5)

# find all markers distinguishing cluster 5 from clusters 0 and 3
cluster5.markers <- FindMarkers(data, 
                                ident.1 = 5, 
                                ident.2 = c(0, 3), 
                                min.pct = 0.25)
head(cluster5.markers,
     n = 5)

# find markers for every cluster compared to all remaining cells, report only the positive ones
data.markers <- FindAllMarkers(data, 
                               only.pos = TRUE, 
                               min.pct = 0.25, 
                               logfc.threshold = 0.25)
data.markers %>%
  group_by(., cluster) %>% 
  top_n(., n = 2, wt = avg_log2FC)

```

## Additional visualizations

```{r}
VlnPlot(data, features = c("MS4A1", "CD79A"))

FeaturePlot(data, features = c("MS4A1", "CD79A"))
```

```{r}
sessionInfo()
```

## THE END.
